Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-07-24 manuscript.tex

In the following, we summarize the recommended theoretical background knowlegde of Markov state modeling for this tutorial.
Then, we address the software requirements to work through the lessons.

@@

In short, the workflow consists of extracting molecular features from the raw data (01) which we transform into a suitable, low-dimensional subspace and discretize (02).
Then, we estimate a maximum likelihood or Bayesian MSM from the discrete trajectories and perform validation tests (03).
We can then analize the stationary and kinetic properties of the MSM (04),
find metastable macrostates and apply transition path theory (TPT) to identify the pathways of conformation change (05),
and compute expectation values for experimental observables (06).

@@

\begin{figure}
\includegraphics{figure_2}
\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a) The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation.
(b) The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
(c) The  sample density projected of the onto the first two independent components of a TICA transformation at lag time $\tau=0.5$ ns shows multiple density maxima and
(d) the timeseries of the first two independent components show rare transition events.
(e) The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for estimating an MSM;
the shaded areas indicate $95\%$ confidence intervals. (f) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of four metastable states accurately predicts the kinetic behavior on longer timescales;
the shaded areas indicate $95\%$ confidence intervals.}
\label{fig:io-to-ck}
\end{figure}

To begin, we start with the raw trajectory data, and we seek to constuct and validate an MSM, which is represented by a transition matrix.

It is initially unknown which features (e.g. distances, contact maps, angles, etc.) are good descriptors of the slow CVs, i.e., are best suited to resolve the long-time dynamics~\cite{NoeClementiReview}.
A systematic way of making this choice has been introduced through the variational principle of conformation dynamics~\cite{noe-vac,nueske-vamk} and recently the more general version of nonequilibrium data, VAMP~\cite{vamp-preprint}.
VAMP provides a score that can be calculated for each feature in order to find good indicators of the long-time dynamics.
This score is computed via cross-validation in order to balance underfitting (too few features) and overfitting (too many features)~\cite{gmrq}.
Here we utilize the VAMP-2 score, which maximizes the kinetic variance of the estimated dynamical operator~\cite{kinetic-maps}.
To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at lag time $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
We note that recently, deep learning approaches for feature selection have been developed that may eventually replace the feature selection step~\cite{vampnet,tae,hernandez-vde}.

@@

Finally, we estimate a (Bayesian) MSM at a the lag time $\tau=0.5$ ns and validate the model using a Chapman-Kolmogorov (CK) test, which compares the right and the left side of the Chapman-Kolmogorov equation
\begin{equation}
\label{eq:ck}
T(k \tau) = T^k(\tau)
\end{equation}
with $T$ being the MSM transition matrix.
To perform this test, we first choose a number of metatstable states appropriate for our system. Then, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$.
As we have identified three slow relaxation timescales in the ITS test,
we perform the CK test on the four metastable states whose kinetics involves these three relaxation timescales.
The CK test shows indeed that our model estimated at lag time tau predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).

@@

\begin{figure}
\includegraphics{figure_3}
\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone:
(a) The reweighted free energy surface projected onto the first two independent components exhibits five minima which
(b) PCCA++ identifies as four metastable states.
(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the two most probable states ($\mathcal{S}_3$, $\mathcal{S}_4$), whereas
(d) the commitor $\mathcal{S}_2\to\mathcal{S}_3$ indicates that the most probable state $\mathcal{S}_4$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_3$.
(e) The Trp-1 SASA shows high variance for all macrostates except $\mathcal{S}_1$ in which the Trp-1 shows the lowest exposure to the solvent.
(f) The Trp-1 SASA auto-correlation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the non-equilibrium condition $\mathcal{S}_1$ (bottom).
(g) Super-imposed cartoon representations of $200$ molecular conformations sampled from each of the four macrostates show that the backbone appears to obstruct the Trp-1 exposure to the solvent in $\mathcal{S}_{(1,2)}$ but less so in $\mathcal{S}_{(3,4)}$.}
\label{fig:msm-analysis}
\end{figure}

@@

To illustrate how to connect an MSM analysis with experiments, we compute the Trp-1 solvent accessible surface area (SASA) using the mdtraj~\cite{mdtraj} package.
The SASA is a reasonable approximation for the instantaneous tryptophan flourescence signal.
The auto correlation function (ACF) of this signal can be measured in spectroscopy experiments.

@@

In this section, we have summarized how to conduct an MSM-based analysis of biomolecular dynamcis data using PyEMMA. For the full analysis, please refer to the first notebook (00). All notebooks as well as detailed installation instructions are available on \githubrepository{}.

>>>>

In the following, we summarize the recommended theoretical background knowledge of Markov state modeling for this tutorial.
Then, we address the software requirements to work through the lessons.

@@

In short, the workflow consists of extracting molecular features from the raw data (01) which we transform into a suitable, low-dimensional subspace and discretize (02).
Then, we estimate a maximum likelihood or Bayesian MSM from the discrete trajectories and perform validation tests (03).
We can then analyze the stationary and kinetic properties of the MSM (04),
find metastable macrostates and apply transition path theory (TPT) to identify the pathways of conformation change (05),
and compute expectation values for experimental observables (06).

@@

\begin{figure}
\includegraphics{figure_2}
\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a) The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation.
(b) The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
(c) The  sample density projected of the onto the first two independent components of a TICA transformation at lag time $\tau=0.5$ ns shows multiple density maxima and
(d) the time series of the first two independent components show rare transition events.
(e) The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for estimating an MSM;
the shaded areas indicate $95\%$ confidence intervals. (f) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of four metastable states accurately predicts the kinetic behavior on longer timescales;
the shaded areas indicate $95\%$ confidence intervals.}
\label{fig:io-to-ck}
\end{figure}

To begin, we start with the raw trajectory data, and we seek to construct and validate an MSM, which is represented by a transition matrix.

It is initially unknown which features (e.g. distances, contact maps, angles, etc.) are good descriptors of the slow CVs, i.e., are best suited to resolve the long-time dynamics~\cite{NoeClementiReview}.
A systematic way of making this choice has been introduced through the variational principle of conformation dynamics~\cite{noe-vac,nueske-vamk} and recently the more general version of non-equilibrium data, VAMP~\cite{vamp-preprint}.
VAMP provides a score that can be calculated for each feature in order to find good indicators of the long-time dynamics.
This score is computed via cross-validation in order to balance underfitting (too few features) and overfitting (too many features)~\cite{gmrq}.
Here we utilize the VAMP-2 score, which maximizes the kinetic variance of the estimated dynamical operator~\cite{kinetic-maps}.
To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at lag time $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
We note that recently, deep learning approaches for feature selection have been developed that may eventually replace the feature selection step~\cite{vampnet,tae,hernandez-vde}.

@@

Finally, we estimate a (Bayesian) MSM at a the lag time $\tau=0.5$ ns and validate the model using a Chapman-Kolmogorov (CK) test, which compares the right and the left side of the Chapman-Kolmogorov equation
\begin{equation}
\label{eq:ck}
T(k \tau) = T^k(\tau)
\end{equation}
with $T$ being the MSM transition matrix.
To perform this test, we first choose a number of metastable states appropriate for our system. Then, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$.
As we have identified three slow relaxation timescales in the ITS test,
we perform the CK test on the four metastable states whose kinetics involves these three relaxation timescales.
The CK test shows indeed that our model estimated at lag time tau predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).

@@

\begin{figure}
\includegraphics{figure_3}
\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone:
(a) The reweighted free energy surface projected onto the first two independent components exhibits five minima which
(b) PCCA++ identifies as four metastable states.
(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the two most probable states ($\mathcal{S}_3$, $\mathcal{S}_4$), whereas
(d) the committor $\mathcal{S}_2\to\mathcal{S}_3$ indicates that the most probable state $\mathcal{S}_4$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_3$.
(e) The Trp-1 SASA shows high variance for all macrostates except $\mathcal{S}_1$ in which the Trp-1 shows the lowest exposure to the solvent.
(f) The Trp-1 SASA auto-correlation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the non-equilibrium condition $\mathcal{S}_1$ (bottom).
(g) Super-imposed cartoon representations of $200$ molecular conformations sampled from each of the four macrostates show that the backbone appears to obstruct the Trp-1 exposure to the solvent in $\mathcal{S}_{(1,2)}$ but less so in $\mathcal{S}_{(3,4)}$.}
\label{fig:msm-analysis}
\end{figure}

@@

To illustrate how to connect an MSM analysis with experiments, we compute the Trp-1 solvent accessible surface area (SASA) using the mdtraj~\cite{mdtraj} package.
The SASA is a reasonable approximation for the instantaneous tryptophan florescence signal.
The auto correlation function (ACF) of this signal can be measured in spectroscopy experiments.

@@

In this section, we have summarized how to conduct an MSM-based analysis of biomolecular dynamics data using PyEMMA. For the full analysis, please refer to the first notebook (00). All notebooks as well as detailed installation instructions are available on \githubrepository{}.
